---
date: "2023-06-23"
date-modified: last-modified
---

# Pipe: magrittr and native {#sec-title}

```{r}
#| label: common.r
#| echo: false
source("_common.R")
```

## Resources
- [magrittr package](https://magrittr.tidyverse.org)
- [Release of R 4.1](https://stat.ethz.ch/pipermail/r-announce/2021/000670.html)
- [The (updated) history of the pipe operator in R](http://adolfoalvarez.cl/blog/2021-09-16-plumbers-chains-and-famous-painters-the-history-of-the-pipe-operator-in-r/)
- [Simpler R coding with pipes > the present and future of the magrittr package -- 5 August 2014](https://www.r-statistics.com/2014/08/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/)
- [Differences between the base R and magrittr pipes](https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/)
- [Isabella Velásquez - Understanding the native R pipe |>](https://ivelasq.rbind.io/blog/understanding-the-r-pipe/)
- [Tim Tiefenbach - Why continue to use magrittr pipe -- 17 October 2022](https://threadreaderapp.com/thread/1582118234220007424.html)

```{r}
#| label: setup
#| message: false
library(dplyr)
library(purrr)
```

## Native pipe {#sec-native-pipe}
- The native pipe was introduced in R 4.1 alongside the use of new anonymous function syntax.
- Pipe the left-hand side into the *first argument* of the right-hand side. In normal usage, it works almost exactly like the `magrittr` pipe.
- In [R 4.2](https://stat.ethz.ch/pipermail/r-announce/2022/000683.html) the native pipe got the `_` syntax to insert the left-hand side into a named argument of the right-hand side function. However, this functionality is not as powerful as in `%>%`, see @sec-magrittr.

The tidyverse style guidelines are moving over to the usage of the native pipe. The native pipe is used for all examples in the [Second Edition of R for Data Science](https://r4ds.hadley.nz). Hadley discusses the move to the native pipe in Documentation in the release notes for [purrr 1.0.0](https://www.tidyverse.org/blog/2022/12/purrr-1-0-0/#documentation). There, he notes the advantages of the greater simplicity of the native pipe and the clarity provided by anonymous functions, see @sec-anonymous.

### Usage {#sec-usage}

The pipe operator is implemented as a syntax transformation.

```{r}
#| label: syntax-transformation
quote(mtcars |> subset(cyl == 4) |> nrow())
```

Like the `magrittr` pipe, the native pipe places the left-hand side into the first argument of the right hand side.

```{r}
#| label: native-pipe
mtcars |> filter(cyl == 4) |> head()
```

There are two ways to place the left-hand side elsewhere in the right-hand side function: anonymous functions and `_` placeholder

### The pipe and the anonymous function {#sec-anonymous}

The pipe and the new shorthand syntax for the anonymous function both debuted with [R 4.1](https://stat.ethz.ch/pipermail/r-announce/2021/000670.html).

```{r}
#| label: anonymous
mtcars |> subset(cyl == 4) |> (\(x) lm(mpg ~ disp, data = x))()
```

Example of using anonymous function to make code more clear taken from [purrr 1.0.0 release](https://www.tidyverse.org/blog/2022/12/purrr-1-0-0/#documentation). This is part of the move away from formula notation (`~`) to anonymous function.

```{r}
#| label: clarity
# Previously
1:5 %>%
  map(~ rnorm(10, .x)) %>%
  map_dbl(mean)

# Use of anonymous function
1:5 |>
  map(\(x) rnorm(10, x)) |>
  map_dbl(mean) 
```

### Underscore placeholder {#sec-underscore}

The `_` placeholder is similar to `magrittr` `.` placeholder. However, the `_` placeholder can only be used *once* and must be used with a *named argument*. See @sec-magrittr for examples of these limitations. Nevertheless, the `_` placeholder does cover most use cases.

```{r}
#| label: underscore
mtcars |> subset(cyl == 4) |> lm(mpg ~ disp, data = _)
```

In [R 4.3](https://stat.ethz.ch/pipermail/r-announce/2023/000691.html) the `_` placeholder can be used with extraction functions `$`, or as the head of a chain of extractions `[`, `[[`, or `@`. However, it cannot be used with `[[` to start the extraction as in case 4 of the magrittr pipe below.

```{r}
#| label: extraction
#| error: true
mtcars |> _$cyl
mtcars |> subset(cyl == 4) |> lm(mpg ~ disp, data = _) |> _$coef[[2]]
```

## The magrittr pipe {#sec-magrittr}

Uses of the `magrittr` `.` notation that are easier or not possible with the native pipe.

1. Use of `.` in unnamed arguments

```{r}
#| label: unnamed-args
c(1:3) %>% paste("No", .)
```

2. Use of multiple `.` in right-hand side

```{r}
#| label: multiple-args
c(1:3) %>% paste(., "No", .)

# Need anonymous function with native pipe
c(1:3) |> (\(x) paste(x, "No", x))()
```

3. Use of `.` in nested functions

```{r}
#| label: nested
iris %>% 
  bind_rows(mutate(., Species = "all")) %>% 
  count(Species)
```

4. Use of `.` with infix operators on left- and right-hand side

```{r}
#| label: infix
mtcars %>% `[[`("cyl")
```
