---
toc: false
number-depth: 0
execute: 
  freeze: true
---

# readr example data {#sec-exdata}

This document creates some toy data to be read in by `readr` as shown in [`readr` notes](01-readr.qmd).

```{r}
#| label: data

my_vctr <- function(x) {
  sample(x, 50, replace = TRUE)
}
my_dates <- function(x) {
  formatC(my_vctr(x), width = 2, flag = "0")
}
years <- 1200:1700
months <- 1:12


set.seed(240)
df <- data.frame(
  a = my_vctr(letters),
  b = my_vctr(c("", "NA", letters[1:5])),
  c = runif(50, 10, 100), 
  d = my_vctr(c("T", "F")),
  e = my_vctr(c("hello", "goodbye", "kook", "gnarly", "none")), 
  f = my_vctr(20:50),
  g = paste0("$", formatC(runif(50, 1000, 200000), digits = 2, big.mark = ",", format = "f")),
  h = paste0(my_vctr(years), my_dates(months), my_dates(1:28)),
  i = paste(my_vctr(years), my_dates(months), my_dates(1:28), sep = "-"),
  j = paste(my_vctr(1:28), my_vctr(month.name), my_vctr(years)),
  k = paste(my_vctr(month.abb), my_vctr(1:28), my_vctr(years)),
  l = my_vctr(10000:500000),
  m = paste0("2023", "-", my_dates(months), "-", my_dates(1:28), " ", my_dates(0:23), ":", my_dates(0:59))
)

head(df)
```

## Column type specifications {#sec-key}
- a: character
- b: character with `NA`s as blank character and "NA"
- c: double
- d: logical
- e: factor with optional `NA` with "none"
- f: integer
- g: number
- h: date: "20230316"
- i: date in locale: "2023-03-16"
- j: date: 16 March 2023
- k: date: Mar 16 2023
- l: time
- m: datetime in locale: "2023-03-16 11:38"

### Write the data frame to `data/`

```{r}
#| label: write.csv
write.csv(df, "data/example.csv", row.names = FALSE)
```
